Large squeezing in multicomponent trapped Bose-Einstein condensates

### 2013-02-25 large_number_squeezing_with_atoms.tex
Minor updates to Bog section
Fix broken IOP/amsmath compatibility.
Add figures showing experimentalist's scaling in 1D, 2D and 3D

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass{iopart}
%\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{ulem}
\usepackage{amssymb}
\usepackage{caption}
\captionsetup{compatibility=false}
\usepackage{subcaption}

@@

We can understand the origin and scaling of this degradation of squeezing by considering the response of the condensate to small fluctuations about the mean field.  We do this using Bogoliubov theory \cite{PethickSmith}, which considers the first order quantum-mechanical fluctuations about the mean field.  To simplify the analysis we consider the case in which $U_{aa} \neq 0$, $U_{bb}=U_{ab} = 0$ and the condensate has a spatially-constant density in a box of side lengths $L_x$, $L_y$ and $L_z$.  In this limit the squeezing $S_a = \Var(N_a)/N_a$ of the $a$ mode at the end of the pulse sequence is (see \ref{appendixBogDerivation})
\begin{align}
  S_a(t_3) &= \frac{\Var[N_a](t_3) + \sum_{\mathbf{k}\neq 0} \frac{1}{2} n_a(\mathbf{k}, t_3) \left\{3 + \cos\left[\Omega(t_3-t_2)\right] + 4 n_a(\mathbf{k}, t_3)\right\}}{N_a(t_3) + \sum_{\mathbf{k}\neq 0} n_a(\mathbf{k}, t_3)}, \label{eqBogoliubovSqueezing}\\
  n_a(\mathbf{k}, t_3) &= \left\{n_a \chi_{aa} \tau_\text{hold} \cos\left[\Omega\left(t_3-t_2\right)\right] \sinc\left(\omega_\mathbf{k} \tau_\text{hold}\right) \right\}^2,\\
  \omega_\mathbf{k} &= \sqrt{\omega^0_\mathbf{k}(\omega^0_\mathbf{k} + 2 \chi_{aa} n_a)}, \\
  \omega^0_\mathbf{k} &= \frac{\hbar \mathbf{k}^2}{2 M},
\end{align}
where $n_a(\mathbf{k}, t) = \expect{\hat{a}^\dagger(\mathbf{k}, t) \hat{a}(\mathbf{k}, t)}$ is the occupation of momentum mode $\mathbf{k}$, $\Var[N_a](t)$ and $N_a(t)$ are the number variance and mean number of the $a$ mode from the two-mode model in Section~\ref{secTwoModeAnalytic}, the sum over $\mathbf{k}$ is a discrete sum over the available non-zero momentum modes $\mathbf{k} = (2\pi/L_x) n_x \hat{\mathbf{x}} + (2\pi/L_y) n_y \hat{\mathbf{y}} + (2\pi/L_z) n_z \hat{\mathbf{z}}$ with $n_x$, $n_y$ and $n_z$ arbitrary integers, $\omega_\mathbf{k}$ is the frequency of the Bogoliubov mode $\mathbf{k}$ and $\omega^0_\mathbf{k}$ is the frequency of a free particle.

@@

@@

&& + c^2 s^2 [ \hat{a}^{\dagger}_2 \hat{a}_2 + \hat{b}^{\dagger}_2 \hat{b}_2 + 4 \hat{a}^{\dagger}_2 \hat{a}_2 \hat{b}^{\dagger}_2 \hat{b}_2 -e^{-2 i \phi} \hat{a}^{\dagger}_2 \hat{a}^{\dagger}_2 \hat{b}_2 \hat{b}_2 -e^{-2 i \phi} \hat{b}^{\dagger}_2 \hat{b}^{\dagger}_2 \hat{a}_2 \hat{a}_2 ] \nonumber \\

@@

&& + c^2 s^2 [ \hat{b}^{\dagger}_2 \hat{b}_2 + \hat{a}^{\dagger}_2 \hat{a}_2 + 4 \hat{b}^{\dagger}_2 \hat{b}_2 \hat{a}^{\dagger}_2 \hat{a}_2 - e^{2 i \phi} \hat{b}^{\dagger}_2 \hat{b}^{\dagger}_2 \hat{a}_2 \hat{a}_2 -e^{2 i \phi} \hat{a}^{\dagger}_2 \hat{a}^{\dagger}_2 \hat{b}_2 \hat{b}_2 ] \nonumber \\

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass{iopart}
  \expandafter\let\csname equation*\endcsname\relax
  \expandafter\let\csname endequation*\endcsname\relax
%\newcommand{\gguide}{{\it Preparing graphics for IOP journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{ulem}
\usepackage{amssymb}
\usepackage{caption}
\captionsetup{compatibility=false}
\usepackage{subcaption}

@@

We can understand the origin and scaling of this degradation of squeezing by considering the response of the condensate to small fluctuations about the mean field.  We do this using Bogoliubov theory \cite{PethickSmith}, which considers the first order quantum-mechanical fluctuations about the mean field.  To simplify the analysis we consider the case in which $U_{aa} \neq 0$, $U_{bb}=U_{ab} = 0$ and the condensate has a spatially-constant density in a box of side lengths $L_x$, $L_y$ and $L_z$.  In this limit the squeezing $S_a = \Var(N_a)/N_a$ of the $a$ mode at the end of the pulse sequence is (a derivation is given in \ref{appendixBogDerivation})
\begin{align}
  S_a(t_3) &= \frac{\Var[N_a](t_3) + \sum_{\mathbf{k}\neq 0} \frac{1}{2} n_a(\mathbf{k}, t_3) \left\{3 + \cos\left[\Omega(t_3-t_2)\right] + 4 n_a(\mathbf{k}, t_3)\right\}}{N_a(t_3) + \sum_{\mathbf{k}\neq 0} n_a(\mathbf{k}, t_3)}, \label{eqBogoliubovSqueezing}\\
  n_a(\mathbf{k}, t_3) &= \left\{n_a \chi_{aa} \tau_\text{hold} \cos\left[\Omega\left(t_3-t_2\right)\right] \sinc\left(\omega_\mathbf{k} \tau_\text{hold}\right) \right\}^2,\\
  \omega_\mathbf{k} &= \sqrt{\omega^0_\mathbf{k}(\omega^0_\mathbf{k} + 2 \chi_{aa} n_a)}, \\
  \omega^0_\mathbf{k} &= \frac{\hbar \mathbf{k}^2}{2 M},
\end{align}
where $n_a(\mathbf{k}, t) = \expect{\hat{a}^\dagger(\mathbf{k}, t) \hat{a}(\mathbf{k}, t)}$ is the occupation of momentum mode $\mathbf{k}$, $\Var[N_a](t)$ and $N_a(t)$ are the number variance and mean number of the $a$ mode from the two-mode model in Section~\ref{secTwoModeAnalytic}, the sum over $\mathbf{k}$ is a discrete sum over the available non-zero momentum modes $\mathbf{k} = (2\pi/L_x) n_x \hat{\mathbf{x}} + (2\pi/L_y) n_y \hat{\mathbf{y}} + (2\pi/L_z) n_z \hat{\mathbf{z}}$ with $n_x$, $n_y$ and $n_z$ arbitrary integers, $\omega_\mathbf{k}$ is the frequency of the Bogoliubov mode $\mathbf{k}$ and $\omega^0_\mathbf{k}$ is the frequency of a free particle.

@@

Some notes on the Bog scaling plots below: Experimentalist's scaling works as follows: V $\rightarrow$ r V; hold time $\rightarrow$ r*(hold time). In order to get enough lattice points to cover the momentum modes up to energy mu, the scaling goes as: 1D lattice$\rightarrow$lattice*$\sqrt{r}$ (1D); lattice$\rightarrow$lattice (2D); lattice$\rightarrow$lattice*$r^{-1/6}$.

The plots show that the experimentalist's scaling increases Bog occupation (i.e. condensate depletion) with volume in 1D, is constant in 2D, and decreases it in 3D. This means squeezing gets worse with volume in 1D, is constant with volume in 2D, and gets better with volume in 3D. We can't actually show squeezing getting better in 3D due to TWA breakdown, but we do show bog mode occupation decreasing.

The jaggedness of the line in 3D is due to the fact that we're only dealing with 10-20 lattice points in each dimension, and the lattice cutoff is often part way between two integer numbers, so we have to round up or down as we change the scaling factor.

Jaggedness of squeezing line in 2D is due to stochastic nature of simulations.

\begin{figure}
  \centering
   \includegraphics[width=7cm]{figures/1D_squeezing_scaling.eps}
\caption{1D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
 \label{fig1D_squeezing_scaling}
\end{figure}

\begin{figure}
  \centering
   \includegraphics[width=7cm]{figures/1D_Bog_mode_occupation_scaling.eps}
\caption{1D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
 \label{fig1D_Bog_mode_occupation_scaling}
\end{figure}

\begin{figure}
  \centering
   \includegraphics[width=7cm]{figures/2D_squeezing_scaling.eps}
\caption{2D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
 \label{fig2D_squeezing_scaling}
\end{figure}

\begin{figure}
  \centering
   \includegraphics[width=7cm]{figures/2D_Bog_mode_occupation_scaling.eps}
\caption{2D experimentalist's scaling. $N=1\times 10^5$, $L=100\mu$m.}
 \label{fig2D_Bog_mode_occupation_scaling}
\end{figure}

\begin{figure}
  \centering
   \includegraphics[width=7cm]{figures/3D_Bog_mode_occupation_scaling.eps}
\caption{3D experimentalist's scaling. $N=1\times 10^4$, $L=20\mu$m.}
 \label{fig3D_Bog_mode_occupation_scaling}
\end{figure}

@@

&& + c^2 s^2 [ \hat{a}^{\dagger}_2 \hat{a}_2 + \hat{b}^{\dagger}_2 \hat{b}_2 + 4 \hat{a}^{\dagger}_2 \hat{a}_2 \hat{b}^{\dagger}_2 \hat{b}_2 -e^{2 i \phi} \hat{a}^{\dagger}_2 \hat{a}^{\dagger}_2 \hat{b}_2 \hat{b}_2 -e^{-2 i \phi} \hat{b}^{\dagger}_2 \hat{b}^{\dagger}_2 \hat{a}_2 \hat{a}_2 ] \nonumber \\

@@

&& + c^2 s^2 [ \hat{b}^{\dagger}_2 \hat{b}_2 + \hat{a}^{\dagger}_2 \hat{a}_2 + 4 \hat{b}^{\dagger}_2 \hat{b}_2 \hat{a}^{\dagger}_2 \hat{a}_2 - e^{-2 i \phi} \hat{b}^{\dagger}_2 \hat{b}^{\dagger}_2 \hat{a}_2 \hat{a}_2 -e^{2 i \phi} \hat{a}^{\dagger}_2 \hat{a}^{\dagger}_2 \hat{b}_2 \hat{b}_2 ] \nonumber \\
